Parametricity for Primitive Nested Types and GADTs

### 2021-06-29 21-lmcs/21-lmcs.tex
5th LMCS revisions

Note that since the body $F$ of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables.  It is in fact
possible to allow ``extra'' $0$-ary functorial variables in the body
of $\mu$-types (functorial variables of higher arity are the real
problem), which would allow the first-order representations of ADTs to
be functorial. However, doing this requires some changes to the
formation rule for $\mu$-types, as well as the delicate threading of
some additional conditions throughout our model construction.  But
since we can always use an ADT's (semantically equivalent)
second-order representation when functoriality is needed, disallowing
such ``extra'' variables does not negatively impact the expressivity
of our calculus. We therefore pursue the simpler syntax here.

@@

It is worth noting that the third component in Equation~(\ref{eq:mu})
is the colimit in $[\rel^k,\rel]$ of third components of relation
transformers, rather than a fixpoint of an endofunctor on
$[\rel^k,\rel]$. There is thus an asymmetry between the first two
components of $\mu T$ and its third component, which reflects the
important conceptual observation that the third component of an
endofunctor on $RT_k$ need not be a functor on all of
$[\rel^k,\rel]$. In particular, although we can define $T_{H,\rho}\,
F$ for a relation transformer $F$ in Definition~\ref{def:rel-sem}
below, it is not clear how we could define it for an arbitrary $F :
[\rel^k,\rel]$. This observation will be especially critical in
Section~\ref{sec:GADTs}.

@@

Intuitively, $\Lan^{\ol{\alpha^0}}_{\ol K} F$ is a syntactic representation
of the left Kan extension of the functor in the variables in
$\ol\alpha$ denoted by $F$ along the functor in the variables in
$\ol\alpha$ denoted by $\ol K$. Using $\Lan$-types, we can therefore
represent the GADT $\mathtt{Seq}$ in our calculus as 
 \begin{align*}
\mathtt{Seq}\,\alpha &\df \left( \mu \phi.\lambda \beta.
\beta + \left(\Lan^{\gamma_1,\gamma_2}_{\gamma_1 \times \gamma_2}
(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. 
\left. + \left(\Lan^{\gamma}_{\mathtt{Nat} \to \gamma} (\mathtt{Nat} \to \phi
\gamma) \right) \beta \right) \alpha
\end{align*}
\noindent
where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat}
\to C$ abbreviates $(\Lan^\alpha_{\mathtt{Nat}} 1)\,C$.  This
representation of $\mathit{Seq}$ corresponds exactly to the rewriting
in Section~\ref{sec:lke}.  As explained in Section~IV.\,D
of~\cite{jp19}, the more general construct $\Lan^{\ol{\alpha^0}}_{\ol
  K} F$ allowing extensions along vectors of functors as depicted
above makes it possible to represent GADTs with two or more type
arguments that depend on one another. Such GADTs cannot be represented
using just unary $\Lan$-types, i.e, $\Lan$-types of the form
$(\Lan^{\ol{\alpha^0}}_K F) A$.

>>>>

Note that since the body $F$ of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables.  It is in fact
possible to allow ``extra'' $0$-ary functorial variables in the body
of $\mu$-types (functorial variables of higher arity are the real
problem), which would allow the first-order representations of ADTs to
be functorial. However, since doing this requires some changes to the
formation rule for $\mu$-types, as well as the delicate threading of
some additional conditions throughout our model construction, we do
not pursue this line of investigation here.
%But since we can always use an ADT's (semantically equivalent)
%second-order representation when functoriality is needed, disallowing
%such ``extra'' variables does not negatively impact the expressivity
%of our calculus. We therefore pursue the simpler syntax here.

@@

It is worth noting that the third component in Equation~(\ref{eq:mu})
is the colimit in $[\rel^k,\rel]$ of third components of relation
transformers, rather than a fixpoint of an endofunctor on
$[\rel^k,\rel]$. There is thus an asymmetry between the first two
components of $\mu T$ and its third component, which reflects the
important conceptual observation that the third component of an
endofunctor on $RT_k$ need not be a functor on all of
$[\rel^k,\rel]$. In particular, although we can define $T_{H,\rho}\,
F$ for a relation transformer $F$ in Definition~\ref{def:rel-sem}
below, it is not clear how we could define it for an arbitrary $F :
[\rel^k,\rel]$.

@@

Intuitively, $\Lan^{\ol{\alpha^0}}_{\ol K} F$ is a syntactic representation
of the left Kan extension of the functor in the variables in
$\ol\alpha$ denoted by $F$ along the functor in the variables in
$\ol\alpha$ denoted by $\ol K$. Using $\Lan$-types, we can therefore
represent the GADT $\mathtt{Seq}$ in our calculus as 
 \begin{align*}
\mathtt{Seq}\,\alpha &\df \left( \mu \phi.\lambda \beta.
\beta + \left(\Lan^{\gamma_1,\gamma_2}_{\gamma_1 \times \gamma_2}
(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. 
\left. + \left(\Lan^{\gamma}_{\mathtt{Nat} \to \gamma} (\mathtt{Nat} \to \phi
\gamma) \right) \beta \right) \alpha
\end{align*}
\noindent
where $\mathtt{Nat} = \mu \alpha. 1 + \alpha$ and $\mathtt{Nat}
\to C$ abbreviates $(\Lan^\emptyset_{\mathtt{Nat}} 1)\,C$.  This
representation of $\mathit{Seq}$ corresponds exactly to the rewriting
in Section~\ref{sec:lke}.  As explained in Section~IV.\,D
of~\cite{jp19}, the more general construct $\Lan^{\ol{\alpha^0}}_{\ol
  K} F$ allowing extensions along vectors of functors as depicted
above makes it possible to represent GADTs with two or more type
arguments that depend on one another. Such GADTs cannot be represented
using just unary $\Lan$-types, i.e, $\Lan$-types of the form
$(\Lan^{\ol{\alpha^0}}_K F) A$.
